MATD02 - Exploration of Finite Geometry

### 2024-11-15 paper/main.tex
Finish proof for theorem 2

%\setlength{\parskip}{\baselineskip}%
%\setlength{\parindent}{0pt}
\usepackage{parskip}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{amsfonts}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{biblatex}

@@

    Finite geometries provide a good introduction to axiomatic, sometimes called synthetic, definitions of geometries
    because they work with finite number of points that can help visualize more abstract concepts.
    \todo{Look over intro again, finite is nice for abstract sets I guess.}

    We'll begin by describing finite geometries using two sets and the set containment relation
    (noting these definitions often differ based on the context.)

@@

    We can now categorize these spaces using axioms, starting with planes.
    Two broad types of plane geometries come from a similar set of axioms which differ in the existence of parallel lines.
    Affine geometry, where parallel lines exist, and projective geometry, where they do not.
    As it turns out, these are quite similar despite projective geometry being simpler to work with\cite[p.~7]{beutelspacher_projective_2000}.

    \subsection{Finite projective planes}

    Let's begin by defining projective spaces axiomatically, with an extra axiom for projective planes,

    \begin{definition}
        A projective space is a geometry such that,
        \begin{itemize}
            \item A1 - For distinct points $p, q$, there is exactly one line $L$ such that $\{p,q\} \subseteq L$.
            We call it the line containing $p, q$, or just the line $pq$.
            \item A2 - Given points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$
            \item A3 - Any line contains at least 3 elements.
            \item A4 - We call this projective space a projective plane if any two lines have at least one point in common, i.e. they intersect
        \end{itemize}
    \end{definition}

    To avoid simple cases, we often also assume that the space has at least two lines.
    The Fano plane is an example of such a projective plane, in fact it is the smallest projective plane,
    containing 7 points and 7 lines, illustrated below:

%   From: https://tex.stackexchange.com/questions/208894/how-might-i-typeset-the-fano-plane-in-latex
    \begin{figure}[h]
        \centering
        \label{fig:fano_plane}
        \tikz[every node/.style={circle, fill, scale=0.5}]
        \draw circle [radius=1] (90:2) -- (210:2) -- (330:2) -- cycle (0,0) node {}
        \foreach \i in {90,210,330}{ (\i:2) node {} -- (\i+180:1) node {} };
    \end{figure}

    It's fairly routine work to show the set of points and lines shown above do in fact satisfy our earlier axioms,
    however more interesting is that this plane has the same number of points and all lines contain three points.
    Does this generalize to all projective spaces?

@@

    In fact, these do generalize, we call the first concept duality,
    in which theorems about lines on a projective plane also have duals regarding points (and vice-versa).
    This generalizes further to higher dimensions with dual spaces.
    The second only makes sense in the finite case, and gives rise to a concept of order. \cite[p. 24]{beutelspacher_projective_2000}

    \begin{theorem}
        Given a finite projective space, there exists a positive integer $q$ such that any line in $L$ contains exactly $q+1$ elements.
        We call q the order of this space.
    \end{theorem}

@@

        Case: $L_1$ intersects $L_2$

@@

        Given points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$
        Given points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$
    \end{proof}

@@

    In fact, when looking at projective geometry through the lens of infinite fields like $\mathbb{R}$,
    we get a definition of projective spaces by looking at one-dimensional vector subspaces of a vector space like $\mathbb{R}^n$.\cite{weisstein_projective_nodate}
    Here, our dimension comes from a concept of coordinates in the underlying vector space, i.e.\ its basis.

    In the finite case\textemdash especially working purely from axioms\textemdash
    we don't have such an obvious idea of dimension.
    We can, however, generalize concepts from vector spaces to get the following definitions.\cite{noauthor_projective_2024,beutelspacher_projective_2000}

@@

    \begin{proof}
        By A4, any two lines $L_1, L_2$ intersect.
        Thus, any nonempty subspace $X$ containing a line $L$
        \todo{Prove.}
    \end{proof}

    This proof highlights the idea behind this geometric dimension definition,
    our chain of subspaces is akin to talking about different objects of increasing dimension in our space.
    That is, points in lines, lines in planes,\ldots, up to $n-1$ dimensional hyperplanes, to the full space itself.

    \subsection{Finite projective spaces of dimension $\geq 3$ }

    As it turns out, order and dimension are powerful tools for classifying finite geometries.
    In fact, we use the notation $PG(n, q)$ to talk about finite projective spaces of dimension $n$ and order $q+1$
    that arise from a finite field of order $q$, written $GF(q)$ (since they are isomorphic up to order.)\todo{Quickly mention what isomorphism means for projective spaces.}

    The Veblen-Young theorem shows finite projective spaces of dimension $n \geq 3$ can always be described using such a field.
    Finite projective planes are much harder to classify, since they can be constructed in more ways.
    Specifically, non-Desarguesian planes.

    \todo{Mention examples? State Desargues' theorem?}

@@

    The difficulty in classifying finite projective planes gives rise to an open question:

    \textit{Is the order of a finite projective plane always a prime power?}

    The number of non-isomorphic planes are known up to order 10: $1, 1, 1, 1, 0, 1, 1, 4, 0$.
    Moreover, the Bruck-Ryser theorem states for $n \equiv 1,2\ (mod\ 4)$ and $n$ is not the sum of two squares,
    then no plane exists of order $n$.\cite{https://oeis.org/A001231}
    \todo{Insert citation into bibtex}

@@

    % References
    \bibliography{references}
    \bibliographystyle{IEEEtran}
\end{document}

>>>>

%\setlength{\parskip}{\baselineskip}%
%\setlength{\parindent}{0pt}
\usepackage{parskip}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{amsfonts}
\usepackage{enumitem}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{wrapfig}

@@

    Finite geometries provide a good introduction to axiomatic, sometimes called synthetic, definitions of geometries
    because they work with finite number of points that can help visualize more abstract concepts.
    \todo{Look over intro again, finite is nice for abstract sets I guess.}
    We'll begin by describing finite geometries using two sets and the set containment relation.\footnote{
        Both the definition for geometries and projective spaces are modified from \cite[p.~1,7]{beutelspacher_projective_2000},
        which uses an arbitrary symmetric and reflexive relation.
        We opt for set containment for its familiarity.}

@@

    We can now categorize these spaces using axioms, starting with planes.
    Two broad types of plane geometries come from a similar set of axioms which differ in the existence of parallel lines.
    Affine geometry, where parallel lines exist, and projective geometry, where they do not.
    As it turns out, these are quite similar despite projective geometry being simpler to work with.

    We start by defining projective spaces and planes axiomatically:

    \begin{definition}
        A projective space is a geometry such that,
        \begin{itemize}[noitemsep]
            \item A1 - For distinct points $p, q$, there is exactly one line $L$ such that $\{p,q\} \subseteq L$.
            We call it the line containing $p, q$, or just the line $pq$.
            \item A2 - Given distinct points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$
            \item A3 - Any line contains at least 3 elements.
            \item A4 - We call this projective space a projective plane if any two lines have at least one point in common, i.e. they intersect
            \item S1 - To avoid simple cases, we add that at least two lines and points exist.
        \end{itemize}
    \end{definition}

%   From: https://tex.stackexchange.com/questions/208894/how-might-i-typeset-the-fano-plane-in-latex
%    \begin{figure}[h]
    \begin{wrapfigure}{r}{0.25\textwidth}
        \centering
        \label{fig:fano_plane}
        \tikz[every node/.style={circle, fill, scale=0.5}]
        \draw circle [radius=1] (90:2) -- (210:2) -- (330:2) -- cycle (0,0) node {}
        \foreach \i in {90,210,330}{ (\i:2) node {} -- (\i+180:1) node {} };
    \end{wrapfigure}

    The Fano plane, shown to the right, is an example of such a projective plane,
    in fact it is the smallest projective plane, containing 7 points and 7 lines.

    It's fairly routine work to show the set of points and lines shown do in fact satisfy our earlier axioms,
    however more interesting is that this plane has the same number of points and all lines contain three points.
    Does this generalize to all projective spaces?

@@

    In fact, these do generalize, we call the first concept duality,
    in which theorems about lines on a projective plane also have duals regarding points (and vice-versa).
    This generalizes further to higher dimensions with dual spaces.
    The second gives rise to a concept of order\cite[p. 24]{beutelspacher_projective_2000}.

    \begin{theorem}
        Given a projective space, there exists a positive integer $q$ such that any line in $L$ contains exactly $q+1$ elements.
        We call q the order of this space.
    \end{theorem}

@@

        \underline{Case 1}: $L_1$ intersects $L_2$

@@

        Given points $a,b,c,d$, if $ab$ intersects $cd$, then $ac$ intersects $bd$

        \underline{Case 2}: $L_1$ does not intersect $L_2$
    \end{proof}

@@

    In fact, when looking at projective geometry through the lens of infinite fields like $\mathbb{R}$,
    we get a definition of projective spaces by looking at one-dimensional vector subspaces of a vector space like $\mathbb{R}^n$ \cite{weisstein_projective_nodate}.
    Here, our dimension comes from a concept of coordinates in the underlying vector space, i.e.\ its basis.

    Working purely from axioms we don't have such an obvious idea of dimension.
    We can, however, generalize concepts from vector spaces to get the following definitions.\footnote{
        The two definitions come from Wikipedia \cite{noauthor_projective_2024},
        I could not find a text referencing them exactly, my assumption is something cited nearby states them in a modified form.
        Nevertheless, I found them fascinating as a demonstration of dimension in an abstract context.
    }

@@

    \begin{proof}
        Let $P$ be the set of points in our plane, noting $|P|\geq 2$ by S1.

        \underline{Part 1}: \textit{The dimension of our plane must be at least 2.}

        Consider any point $a \in P$, $\{ a \}$ is a subspace of $P$ vacuously since it doesn't contain two points.
        Now, given any other point $b$, $ab$ is the only line that passes through two points of $ab$ by A1.
        So $ab$ is a subspace since $ab \subseteq ab$.
        Finally, $P$ has at least two lines by S1, so $ab \subset P$.
        So we have a chain of subspaces $\emptyset \subset \{a\} \subset ab \subset P$, thus our geometric dimension is at least 2.

        \underline{Part 2}: \textit{The dimension of our plane must be at most 2.}

        In a maximal chain of subspaces, any subspace $X_i$ must have that there does not exist a subspace $X$ for which $X_{i-1} \subset X \subset X_i$.
        Since any nonempty subspace $X$ must contain a point $a$ and $\{a\}$ is a subspace (as shown earlier),
        with no sets existing between $\emptyset$ and $\{a\}$, we must have in our chain that $X_0 = \{a\}$.

        Similarly, since $X_0 \subset X_1$, $X_1$ contains at least another point $b$, and again the line $ab$ is a subspace as shown earlier.
        Any strict subset of $ab \neq \emptyset, \{a\}$ would not contain the line $ab$ in full and thus cannot be a subspace.
        So it must be that $X_1 = ab$.

        Now, since $X_1 \subset X_2$, we must have a distinct $c \in X_2$ such that $c \notin X_2 = ab$.
        So we have three distinct points $a,b,c$ which form three distinct lines by our choice of $c$.
        We'll show any other line in our space must also be in $X_2$.
        Let $x,y \in P$ be arbitrary points, wlog we can say $y \in ab$ since by A4, $xy$ must intersect $ab$ at some point.
        Again by A4, let $xy$ intersect $bc$ at $z_1$, and $ac$ at $z_2$.

        \begin{figure}[h]
            \label{fig:triangle}
            \centering
            \begin{tikzpicture}
                % Draw the lines
                \draw[thick] (-1,0) -- (0,1.732);% node[right] {$L_1$};
                \draw[thick] (1,0) -- (0,1.732);% node[right] {$L_2$};
                \draw[thick] (-1,0) -- (1,0);
                % Draw the intersection point
                \fill (-1,0) circle (2pt) node[below] {$a$};
                \fill (0,1.732) circle (2pt) node[above] {$b$};
                \fill (1,0) circle (2pt) node[below] {$c$};

                \fill (-1.5,1.3) circle (2pt) node[above] {$x$};
                \fill (-0.5,0.85) circle (2pt) node[below] {$y$};
                \fill (0.86346, 0.23644) circle (2pt) node[right] {$z$};
                \draw[thick] (-1.5,1.3) -- (0.86346, 0.23644);
            \end{tikzpicture}
            \caption{Intuitively, regardless of our choice of $x$, since $xy$ also intersects $bc$, and $ac$, we'll have a distinct point $z$ on one of the two lines.}
        \end{figure}

        Suppose $y = z_1 = z_2$, then we have that all three lines $ab,bc,ac$ share one point $y$,
        but then $\{b, y\} \subseteq ab$ and $\{b, y\} \subseteq bc$, so $ab = by = bc$ by A1,
        since $by$ is the unique line passing through $b, y$.
        But this contradicts $ab, bc$ being distinct lines.
        Thus, we must have that $y \neq z$ where $z$ is one of $z_1$ or $z_2$.

        Finally, the line $xy$ contains points $y,z \in X_2$, and since $X_2$ is a subspace we must have that $xy \in X_2$
        So, $X_2$ = $P$ since any two points in $P$ form a line (containing them) that is fully contained in $X_2$.
        Thus, the dimension is at most 2.
    \end{proof}

    This proof highlights the idea behind this geometric dimension definition,
    our chain of subspaces is akin to talking about different objects of increasing dimension in our space.
    That is, points in lines, lines in planes \ldots up to $n-1$ dimensional hyperplanes, to the full space itself.

@@

    As it turns out, order and dimension are powerful tools for classifying finite geometries.
    In fact, we use the notation $PG(n, q)$ to talk about finite projective spaces of dimension $n$ and order $q+1$
    that arise from a finite field of order $q$, written $GF(q)$ (since they are isomorphic up to order.)\cite[p.~xv]{hirschfeld_general_2016}

    The Veblen-Young theorem shows finite projective spaces of dimension $n \geq 3$ can always be described using such a field.
    However, finite projective planes are much harder to classify, since they can be constructed in more ways.
    Specifically, non-Desarguesian planes.
    The difficulty in classifying finite projective planes gives rise to an open question:
    \textit{Is the order of a finite projective plane always a prime power?}
    \todo{Mention examples? State Desargues' theorem?}

    The number of non-isomorphic finite projective planes is only known up to order 10: $1, 1, 1, 1, 0, 1, 1, 4, 0$
    due to the computational complexity of verifying this using brute force\cite{lam_computer_1991,lam_non-existence_1989}.
    Moreover, the closest we have come to a proof of this conjecture is the Bruck-Ryser theorem, which states if $n \equiv 1,2\ (mod\ 4)$ and $n$ is not the sum of two squares,
    then no plane exists of order $n$.

@@

    % References
    \newpage
    \bibliography{references}
    \bibliographystyle{IEEEtran}
\end{document}
